American Indian and Alaska Native child welfare system contact across U.S. States: magnitudes and mechanisms

### 2021-08-22 state_ms.Rmd
add state ms

---
title: "Mechanisms of AIAN Inequality in Child Welfare Across U.S. States."
author: "Frank Edwards and Theresa Rocha Beardall"
date: "December, 2020"
bibliography: "casey_rpt.bib"
link-citations: true
csl: "vancouver-brackets-no-et-al.csl"
abstract: "American Indian and Alaska Native children are separated from their families by state child welfare agencies at exceptionally high rates. This study connects contemporary trends in Native family separation to histories of Indian child removal, and provides insights into the geographies and institutional sites where inequalities emerge. We find that the total number of AIAN children in foster care or adoptive homes in states with large Native populations has increased since the passage of the Indian Child Welfare Act (1978). We find that risks of child welfare system contact are highest for AIAN infants, and that risk is highly variable across states. Our estimates show that in high risk states at 2014 - 2018 levels of risk, more than half of AIAN children will ever be investigated by a child welfare agency, more than one in five will experience a substantiated maltreatment case, and more than one in five will ever enter foster care. We further find that child welfare agency case processing exacerbates inequality. AIAN children are more likely than white children to enter foster care, conditional on experiencing a substantiated maltreatment case, than are white children. These exceptionally high levels of risk indicate that the crisis of Indian family separation is ongoing. For AIAN children in states like Minnesota, Alaska, South Dakota, North Dakota, and Oklahoma, contact with the child welfare system is a routine part of growing up."
output: bookdown::pdf_document2
---

@@

Despite growing attention to threats against the health, safety, and wellbeing of tribal communities in recent years, the child welfare system remains a confounding and persistently complex source of harm in the lives of Native families throughout the U.S. These harms stem from institutional acts of violence waged against Native Peoples by the U.S. federal government. This began with early Americans who were eager to physically remove Native Peoples from their homelands in order to establish commerce, statehood, and homes for non-Native families. Often, the intention of such violence was to break up Native families and assimilate them into the nation’s population. Boarding schools and adoption programs carried the U.S. mission of assimilating Native children well into the 20th century.

Family separation continues to jeopardize and threaten the autonomy and wellbeing of Native families, and as a result, the very existence of tribal communities. These threats remain despite the passage of the landmark Indian Child Welfare Act (ICWA) of 1978, a law that was implemented to address and end generations of abuse, mistreatment, and forcible removal of Native children from their homes.

In the following study, we describe the scale of the ongoing crisis of family separation by 1) evaluating the magnitude of change of family separation since ICWA, 2) identifying geographic variation in child welfare inequality, and 3) identifying the stages in case processing where inequality is accelerated. This study provides new empirical evidence about where and how Native families are being pushed into the child welfare system, and situates these findings in the long historical legacy of Native family separation.

@@

This study relies on a number of data sources. First, we compile and digitize statistics collected by the Association for American Indian Affairs during the late 1960s and early to mid 1970s in an effort to document the breadth and depth of the Indian family separation crisis @ninety-fifth_congress_indian_1977. To capture the contemporary exposure of AIAN children to the child welfare system, we rely on federal data systems collected by the U.S. Children's Bureau. We use the 2014 - 2018 National Child Abuse and Neglect Data System (NCANDS) child files @u_s_department_of_health_and_human_services__administration_on_children_youth_and_families_national_2018, the 2014 - 2018 Adoption and Foster Care Analysis and Reporting System (AFCARS) foster care files @u_s_department_of_health_and_human_services__administration_on_children_youth_and_families_adoption_2018 and the 2010 - 2018 AFCARS adoption files @u_s_department_of_health_and_human_services__administration_on_children_youth_and_families_adoption_2018-1. We rely on the U.S. Census Bureau's Population Estimates Program (PEP) for contemporary population estimates @us_census_bureau_age_2020.

## Historical child welfare system data

The Association on American Indian Affairs submitted a report titled "Indian Child Welfare Statistical Summary, July 1976" in support of the Indian Child Welfare Act (1978) during congressional hearings, included as Appendix G in hearings convened by the Select Committee on Indian Affairs on August 4, 1977. The results presented by AAIA were the result of multiple waves of surveys sent by AAIA to federal, state, local, and private service providers in 19 states with large AIAN populations (Alaska, Arizona, California, Idaho, Maine, Michigan, Minnesota, Montana, Nevada, New Mexico, New York, North Dakota, Oklahoma, Oregon, South Dakota, Utah, Washington, Wisconsin, and Wyoming). AAIA compiled statistics on the characteristics of both Native and non-Native children in foster care, in adoption, or in boarding programs for each of these states.

These data consistently provide point-in-time estimates of the numbers of Native and non-Native children in foster care, adoptions, or in Bureau of Indian Affairs boarding schools. During this period, some states did not collect information on race/ethnicity for children in foster care. When this was the case, AAIA imputed foster care caseload totals for Native and non-Native children by multiplying the total volume of children in foster care by the proportion of the population that was Native and non-Native respectively. This approach (as AAIA acknowledges), likely produced conservative estimates of AIAN foster care caseloads in these states, as Native children were more likely than non-Native children to enter foster care in these states during this period.

These landmark data provided a first close-to-comprehensive national picture of the scale of Indian family separation by combining federal, state, and local data sources, and data provided much of the quantitative empirical foundation for the landmark reforms implemented in the Indian Child Welfare Act. We use them here to measure change in the exposure of AIAN families to foster care and adoption since the passage of ICWA.

## Contemporary child welfare system data

@@

The AFCARS adoption file provides detailed information on all children newly adopted with state or local child welfare agency involvement in the U.S. each year. To provide point-in-time estimates of total numbers of children in adoptive homes, we aggregate data from 2010 - 2019, then evaluate the number of both Native and non-Native children in adoptive homes who would have been 21 years old or younger in 2019. We use the 21-year-old threshold for comparability with data collected by AAIA.

## Population data

@@

We are not able to leverage Census population data directly to make comparisons of the incidence rate of family separation between the 1970s and 2010s. Changes in the enumeration of the AIAN population between 1976 and 2019 make such comparisons impractical. For example, the U.S. Census counted approximately 1.4 million American Indians and Alaska Natives in the 1980 decennial count, a 70 percent increase over the 1970 decennial census. Such dramatic shifts are largely explained by the Census changing its methodology for determining race/ethnicity of a respondent. Until 1970, the Census worker reported a subject's race/ethnicity based on worker observations. In 1980, the Census changed to self-identification of race/ethnicity @lujan_american_2014. In 1990, the Census counted about 2 million American Indians and Alaska Natives, an approximately 40 percent increase from the prior Census. By 2000, the Census reported 4.2 million American Indians and Alaska Natives living in the United States, more than double the 2000 population count. Such changes are not driven by population processes alone, but instead are an artifact of shifting methods of enumerating the racial and ethnic identities of U.S. residents by the Census Bureau. This makes direct comparisons of historical AIAN population counts to contemporary AIAN population counts problematic.

@@

## Historical comparisons

We directly compare counts of children experiencing family separation through adoption and foster care in the states surveyed by AAIA in 1976 to counts of children in foster care or adoption in 2019. Because historical and contemporary AIAN population estimates are incompatible, we do not provide estimates as per capita rates. AAIA reports point-in-time counts of children in foster care or adoption by state. For comparability, we compute point-in-time contemporary foster care caseloads by counting the number of children in foster care at the end of the reporting year in 2019, and adoption caseloads by using annual counts of new adoptions from 2010-2019 and estimating the number of children concurrently in adoptive homes under the age of 21 (for comparability with AAIA measures, which used 21 as the age threshold for inclusion). We compare counts of AIAN and non-AIAN children in each system for states included in the AAIA data at these two time points, then compare aggregate totals of children in some form of state-involved family separation (inclusive of Bureau of Indian Affairs boarding schools) across these two time periods.

@@

## Change Since ICWA

```{r icwa-setup}

### ICWA DATA NAT

icwa_nat<-icwa_fc %>%
  group_by(race_ethn) %>% 
  summarize(fc = sum(fc, na.rm=T),
            adopted = sum(adopted, na.rm=T),
            boarding = sum(boarding, na.rm=T))

##

afcars_caseload19<-read_csv("./data/afcars19_caseload.csv")
afcars_caseload19<-afcars_caseload19 %>% 
  mutate(race_ethn = ifelse(AMIAKN==1,
                            "AIAN",
                            "other")) %>% 
  rename(state = St,
         st_fips = STATE,
         fc = caseload) %>% 
  filter(st_fips!=72) %>% 
  select(st_fips, race_ethn, fc) %>% 
  filter(!is.na(race_ethn))

### join with afcars adoption data
afcars_caseload<-afcars_caseload19 %>% 
  left_join(afcars_adopt_caseload_19 %>% 
  mutate(race_ethn = ifelse(race_ethn=="AIAN",  
                            race_ethn,
                            "other")) %>% 
  mutate(n_non_preferred = ifelse(race_ethn=="AIAN",
                                  n - n_kin_or_aian,
                                  n - n_kin),
         adopt_non_preferred = n_non_preferred / n) %>% 
    rename(st_fips = STATE,
           state = St,
           adopted = n) %>% 
  select(state, race_ethn, adopted, adopt_non_preferred)) %>% 
  mutate(state,  race_ethn, fc, adopted) %>% 
  mutate(period = "2019")

### state comparisons
icwa_delta<-icwa_fc %>% 
  mutate(race_ethn = ifelse(race_ethn=="White",
                            "other",
                            race_ethn)) %>% 
  bind_rows(afcars_caseload %>% 
              select(state, period, race_ethn, fc, adopted)) %>% 
  filter(state%in%icwa_fc$state)

### make comparable national totals
### can't compare per cap rates
### pop denominators aren't comparable

icwa_delta_pct<-
  icwa_delta %>% 
  pivot_wider(id_cols = c(state, race_ethn),
              names_from = period, 
              values_from = c(fc, adopted)) %>% 
  mutate(delta_fc = (fc_2019 / fc_1976 -1) * 100,
         delta_adopt = (adopted_2019/adopted_1976 -1)*100) %>% 
  select(state, race_ethn, delta_fc, delta_adopt) %>% 
  pivot_longer(cols = delta_fc:delta_adopt) %>% 
  mutate(name = ifelse(
    name=="delta_fc",
    "Foster care",
    "Adoption"
  ))

### remove adoption states where we don't have AIAN ICWA numbers
index<-icwa_delta_pct %>% 
  filter(is.na(value))

icwa_delta_pct <- icwa_delta_pct %>% 
  mutate(value = ifelse(state%in%index$state & name == "Adoption",
         NA,
         value))

icwa_nat<-icwa_delta %>% 
  filter(!state%in%index$state) %>% 
  group_by(period, race_ethn) %>% 
  summarise(fc = sum(fc),
            adopted = sum(adopted))

### plus boarding
total_inst<-icwa_delta %>% 
  group_by(period, race_ethn) %>% 
  summarise(fc = sum(fc),
            adopted = sum(adopted, na.rm=T)) %>% 
  mutate(total = fc + adopted)

icwa_nat_na<-icwa_delta %>% 
  group_by(period, race_ethn) %>% 
  summarise(fc = sum(fc, na.rm=T),
            adopted = sum(adopted, na.rm=T))
```

We compare data collected by the Association on American Indian Affairs between 1973 and 1976 on American Indian and Alaska Native children in foster care and adoption to contemporary data on AIAN children in foster care and adoption collected through the Adoption and Foster Care Analysis and Reporting System between 2010 and 2019. Figure \@ref(fig:icwa-fc-counts) shows the total numbers of children in foster care as reported by AAIA in the mid-1970s and in AFCARS in 2019. There were `r icwa_nat_na$fc[1]` Native children in foster care in the 19 states for which AAIA collected data in the mid-1970s. In 2019, there were `r icwa_nat_na$fc[3]` Native children in foster care in these states, an increase of `r round(((icwa_nat_na$fc[3] / icwa_nat_na$fc[1]) - 1) * 100,0)` percent. For non-Native children in these 19 states, caseloads increased by `r round(((icwa_nat_na$fc[4] / icwa_nat_na$fc[2]) - 1) * 100,0)` percent over the same period.

```{r icwa-fc-counts, fig.cap = "Children in foster care on reporting date (caseload) by state and period. Data from AAIA surveys and AFCARS"}
ggplot(icwa_delta %>% 
         mutate(race_ethn = 
                  ifelse(race_ethn=="other",
                         "Non-AIAN",
                         race_ethn)),
       aes(y = state, x = fc, fill = period)) + 
  geom_col(position = position_dodge2(reverse = T)) + 
  facet_wrap(~race_ethn, scales = "free") + 
  scale_fill_brewer(palette = "Dark2") + 
  labs(y = "",
       x = "Children in foster care",
       fill = "") 
```

Figure \@ref(fig:icwa-adopt-counts) displays counts of children 21 years of age or younger in state-involved adoptions for both the mid-1970s and 2019. Note that 6 of the 19 included states lack complete data on AIAN children in adoption in the AAIA data (Idaho, Maine, Nevada, New Mexico, New York, and Wyoming). For the 13 states with complete adoption data, AAIA estimated that there were `r icwa_nat$adopted[1]` Native children in state-involved adoptions in the mid-1970s. In these same 13 states, there were `r icwa_nat$adopted[3]` AIAN children in adoptions in 2019, an increase of `r round(((icwa_nat_na$adopted[3] / icwa_nat_na$adopted[1]) - 1) * 100,0)` percent. For non-native children in these states, there was a `r round(abs(((icwa_nat_na$adopted[2] / icwa_nat_na$adopted[4]) - 1) * 100),0)` percent reduction in the numbers of adopted children over the same period.

```{r icwa-adopt-counts, fig.cap = "Children 21 and under in state-involved adoptions on reporting date by state and period. Data from AAIA surveys and AFCARS"}
  ggplot(icwa_delta %>% 
           mutate(race_ethn = 
                    ifelse(race_ethn=="other",
                           "Non-AIAN",
                           race_ethn)),
         aes(y = state, x = adopted, fill = period)) + 
    geom_col(position = position_dodge2(reverse = T)) + 
    facet_wrap(~race_ethn, scales = "free") + 
    scale_fill_brewer(palette = "Dark2") + 
    labs(y = "",
         x = "Children in adoption",
         fill = "")

#### make state-level proportional change over time 
### for both outcomes
```

Figure \@ref(fig:icwa-pct) shows the relative change in adoption and foster care caseloads between the mid-1970s and 2019 for these states. Of the 19 states, 14 saw growth in the numbers of Native children in foster care since the passage of ICWA. Of the 13 states with complete adoption data, 8 had more Native children in foster care in 2019 than in the mid-1970s prior to the passage of ICWA. For non-Native children, foster care caseloads were higher in 16 of 19 states and adoption caseloads were higher in 6 of the 13 states with complete data. Oklahoma saw the largest absolute growth in Native foster care caseloads. In the mid-1970s, there were `r icwa_delta %>% filter(state=="OK", race_ethn=="AIAN", period == 1976) %>% select(fc) ` Native children in foster care in Oklahoma. In 2019, there were `r icwa_delta %>% filter(state=="OK", race_ethn=="AIAN", period == 2019) %>% select(fc) ` Native children in foster care in Oklahoma, a growth of nearly `r round(icwa_delta_pct %>% filter(state=="OK", name == "Foster care", race_ethn == "AIAN") %>% select(value), 0)` percent.

```{r icwa-pct, fig.cap = "Percent change in adoption and foster care caseloads between 1976 and 2019. Data from AAIA surveys and AFCARS"}

ggplot(icwa_delta_pct %>% 
         mutate(race_ethn =
                  ifelse(race_ethn == "AIAN",
                         race_ethn,
                         "Non-AIAN")),
       aes(x = value,
           y = state,
           fill = name)) + 
  geom_col(position = position_dodge2(reverse = T)) + 
  labs(y = "") +
  geom_vline(xintercept = 0, lty = 2) +
  scale_fill_brewer(palette = "Dark2") + 
  facet_wrap(~race_ethn, scales = "free") + 
  labs(fill = "", 
       x = "Percent increase, 1976 to 2019")
```

```{r icwa-growth}
#COMPUTE GROWTH IN AIAN - GROWTH IN NONAIAN FOR WHICH GREW FASTER.
icwa_delta_aian<-icwa_delta_pct %>% 
  filter(race_ethn=="AIAN") %>% 
  select(-race_ethn) %>% 
  rename(aian_value = value)
icwa_delta_non<-icwa_delta_pct %>% 
  filter(race_ethn!="AIAN") %>% 
  select(-race_ethn) %>% 
  rename(non_aian_value = value)

icwa_pct_diff<-icwa_delta_aian %>% 
  left_join(icwa_delta_non) %>% 
  mutate(diff = aian_value - non_aian_value)

tab_higher<-icwa_pct_diff %>% 
  group_by(name) %>% 
  summarise(aian_higher = sum(diff>0, na.rm=T))
```

Foster care caseloads grew in many states for both non-AIAN and AIAN children. Of the 19 states included here, 12 saw higher growth in AIAN foster care caseloads than they saw in non-AIAN foster care caseloads. Adoption caseloads also grew faster for AIAN children in these states than they did for non-AIAN children. In 7 of 13 states with complete data, AIAN adoption caseloads grew faster between the mid-1970s and 2019 than did non-AIAN adoption caseloads.

Of course, these data do not provide a complete comparative picture of Indian family separation over time. Until the late 20th century, American Indian and Alaska Native children were routinely removed from their families and sent to residential boarding schools. Approximately 25000 Native children were in Bureau of Indian Affairs boarding schools in 1973. This total is larger than the total number of Native children in foster care or adoption recorded by AAIA in 1976. There were more AIAN children in boarding schools in 1973 than were in foster care in 2019. While the scale of separation of Native children from their families through foster care and adoption has increased over time in these states, the absolute scale of the separation of Native children from their families has declined nationally with the end of the boarding school era.

@@

```{r nat-inv-age, fig.cap = "Age specific risk of experiencing first CPS investigation by state, 2014 - 2018 risk levels"}
ggplot(investigation_tables %>% 
         group_by(state, race_ethn, age) %>% 
         summarise(qmax = max(q), qmin=min(q), q = mean(q)),
       aes(x = age, y = q, ymin = qmin, ymax = qmax,
           color = race_ethn, fill = race_ethn)) + 
  geom_line() + 
  geom_ribbon(alpha = 0.5, color = NA) + 
  facet_geo(~state) + 
  labs(color = "", fill = "") + 
  theme_minimal() + 
  labs(x = "Age", y = "Probability") + 
  theme(strip.text.x = element_text(size = 7),
        axis.text.x = element_text(size = 6),
        axis.text.y = element_text(size = 4)) + 
  scale_y_continuous(labels = scales::number_format(accuracy = 0.01)) + 
  scale_fill_brewer(palette = "Dark2")+
  scale_color_brewer(palette = "Dark2")

@@

```{r nat-fc-age, fig.cap = "Age specific risk of entering foster care, 2014 - 2018 risk levels"}
ggplot(fc_tables %>% 
         group_by(state, race_ethn, age) %>% 
         summarise(qmax = max(q), qmin=min(q), q = mean(q)),
       aes(x = age, y = q, ymin = qmin, ymax = qmax,
           color = race_ethn, fill = race_ethn)) + 
  geom_line() + 
  geom_ribbon(alpha = 0.5, color = NA) + 
  facet_geo(~state) + 
  labs(color = "", fill = "") + 
  theme_minimal() + 
  labs(x = "Age", y = "Probability") + 
  theme(strip.text.x = element_text(size = 7),
        axis.text.x = element_text(size = 6),
        axis.text.y = element_text(size = 4)) + 
  scale_y_continuous(labels = scales::number_format(accuracy = 0.01)) + 
  scale_fill_brewer(palette = "Dark2")+
  scale_color_brewer(palette = "Dark2")
```

@@

ggplot(ncands_sub_inv_tab %>% 
  group_by(age, race_ethn, state) %>% 
    summarise(qmin = min(sub_inv),
              qmax = max(sub_inv),
              q = mean(sub_inv)),
  aes(x = age, y = q, color = race_ethn,
      ymin = qmin, ymax = qmax, fill = race_ethn)) +
  geom_line() + 
  geom_ribbon(alpha = 0.5,
              color = NA) + 
  labs(fill = "", color = "",
       y = "Probability", x = "Age") + 
  facet_geo(~state) +
  scale_fill_brewer(palette = "Dark2")+
  scale_color_brewer(palette = "Dark2")+ 
  theme(axis.text.y = element_text(size = 5),
        axis.text.x = element_text(size = 5))
```

@@

ggplot(fc_cond_inv_tabs %>% 
         group_by(age, race_ethn, state) %>% 
         summarise(qmin = min(fc_cond_inv),
                   qmax = max(fc_cond_inv),
                   q = mean(fc_cond_inv)) %>% 
         filter(!(is.na(q)),
                q<1),
       aes(x = age, y = q, color = race_ethn,
           ymin = qmin, ymax = qmax, fill = race_ethn)) +
  geom_line() + 
  geom_ribbon(alpha = 0.5,
              color = NA) + 
  labs(fill = "", color = "",
       y = "Probability", x = "Age") + 
  facet_geo(~state) +
  scale_fill_brewer(palette = "Dark2")+
  scale_color_brewer(palette = "Dark2")+ 
  theme(axis.text.y = element_text(size = 5),
        axis.text.x = element_text(size = 5))
```

@@

ggplot(fc_cond_sub_tabs %>% 
         group_by(age, race_ethn, state) %>% 
         summarise(qmin = min(fc_cond_sub),
                   qmax = max(fc_cond_sub),
                   q = mean(fc_cond_sub)) %>% 
         filter(!(is.na(q)),
                q<1),
       aes(x = age, y = q, color = race_ethn,
           ymin = qmin, ymax = qmax, fill = race_ethn)) +
  geom_line() + 
  geom_ribbon(alpha = 0.5,
              color = NA) + 
  labs(fill = "", color = "",
       y = "Probability", x = "Age") + 
  facet_geo(~state) +
  scale_fill_brewer(palette = "Dark2")+
  scale_color_brewer(palette = "Dark2")+ 
  theme(axis.text.y = element_text(size = 5),
        axis.text.x = element_text(size = 5)) + 
  coord_cartesian(ylim = c(0,1))
```

@@

ggplot(non_icwa_cond_fc %>% 
         filter(race_ethn=="AIAN") %>% 
         group_by(age, state) %>% 
         summarise(qmin = min(non_icwa_fc),
                   qmax = max(non_icwa_fc),
                   q = mean(non_icwa_fc)) %>% 
         filter(!(is.na(q)),
                q<=1, qmax<=1),
       aes(x = age, y = q,
           ymin = qmin, ymax = qmax)) + 
  geom_line() +
  geom_ribbon(alpha = 0.5, color = NA) + 
  facet_geo(~state) + 
  labs(x = "Age", y = "Proportion") + 
  theme(axis.text.y = element_text(size = 5),
        axis.text.x = element_text(size = 5))
```

@@

This report provides two sets of empirical findings.

First, we examine data on the breadth of the separation of AIAN children prior to the passage of the Indian Child Welfare Act, and compare these data to contemporary AIAN family separation through the child welfare system. We find that, while the overall scale of AIAN family separation has decreased since ICWA due to the closing of BIA boarding schools, the scale of AIAN family separation through fostering and adoption has increased in the states for which we have comparable data. In some states, like Oklahoma, Washington, California, and Alaska, the frequency of AIAN foster care and adoption has increased dramatically since the mid-1970s.

However, these results should be interpreted with several limitations in mind. Because of changes in the ways that the U.S. Census tabulates AIAN populations, and because of cultural shifts in how AIAN people identify themselves when asked questions about racial identity on surveys like the Census @lujan_american_2014, estimates of the size of the AIAN child population are not comparable between 1976 and 2014 - 2018. As such, we cannot directly compare population adjusted incidence rates, only raw counts. Nonetheless, the magnitude of growth in the frequency of AIAN foster care and adoption in many states has outpaced the kinds of growth we would expect from demographic processes alone, and suggests that the scope of AIAN fostering and adoption has indeed grown over this period. Future research should explore methods for normalizing population counts across these two periods.

We should also consider AAIA's counts of Native children in foster care and adoption in the pre-ICWA period to be conservative. AAIA relied on agency surveys to count AIAN children that had been separated from their families, and limited survey non-response negatively biases these counts. Additionally, many state and local agencies did not collect data on a child's racial or ethnic identity during this period. In some states, AAIA imputed the total number of AIAN children in foster care and adoption by applying the state or local population composition to the total count of children in foster care or adoption. Assuming any disparity in placement (a reasonable assumption), these imputations would be negatively biased, as AAIA suggests in their report. As such, the pre-ICWA placement numbers are likely conservative estimates of the scale of AIAN family separation in 1976. Because there is likely no such bias in contemporary numbers, estimates of proportional change (as displayed in Figure \@ref(fig:icwa-pct)) may overstate the magnitude of change since ICWA.

Second, we examine data on the contemporary contact of AIAN children with the child welfare system at five distinct stages of case processing: investigations, substantiations, removals, placement settings, and termination of parental rights. We find that AIAN children are more likely than white children to be placed into foster care, even when rates of investigation and substantiation of cases are similar to those for white children. We find that inequalities in case outcomes vary dramatically across regions of the country. Inequalities for AIAN children are typically low in the South and Northeast, and typically very high in the Midwest, Great Plains, Mountain West, and Northwest. While some states, such as Alaska, Minnesota, and North and South Dakota exhibit high inequalities in risks of investigation for AIAN children, we find that inequalities in family separation are typically a function of later case processing decisions. In many places, decisions made once a child is in the system, are critical in exacerbating the unequal exposure of AIAN children to family separation.

>>>>

---
title: "American Indian and Alaska Native child welfare system contact across U.S. States: magnitudes and mechanisms"
author: "Frank Edwards and Theresa Rocha Beardall"
bibliography: "lib.bib"
link-citations: true
#csl: "vancouver-brackets-no-et-al.csl"
abstract: "Following a long history of forced assimilation through family separation, American Indian and Alaska Native children continue to be separated from their families by state child welfare agencies at exceptionally high rates. This study provides insights into the geographies and institutional sites where inequalities emerge. We find that risks of child welfare system contact are highest for AIAN infants, and that risk is highly variable across states. Our estimates show that in high risk states at 2014 - 2018 levels of risk, more than half of AIAN children will ever be investigated by a child welfare agency, more than one in five will experience a substantiated maltreatment case, and more than one in five will ever enter foster care. We further find that child welfare agency case processing exacerbates inequality. AIAN children are more likely than white children to enter foster care, conditional on experiencing a substantiated maltreatment case, than are white children. These exceptionally high levels of risk indicate that the crisis of Indian family separation is ongoing. For AIAN children in states like Minnesota, Alaska, South Dakota, North Dakota, and Oklahoma, contact with the child welfare system is a routine part of growing up."
output: bookdown::pdf_document2
# output: word_document
toc: false
---

@@

The child welfare system remains a confounding and persistently complex source of harm in the lives of Native families throughout the United States [@haight_scoping_2018; @dettlaff_racial_2021]. These harms stem from institutional acts of violence waged against Native Peoples by the U.S. government [@glenn_settler_2015; @wolfe_settler_2006]. This began with early Americans who were eager to physically remove Native Peoples from their homelands in order to establish commerce, statehood, and homes for non-Native families. During the late 19th through the mid-20th century, federal, state, and religious organizations worked to aggressively break up Native families and assimilate them into the nation’s white population through boarding schools, fostering, and adoption [@adams_education_1995; @trennert_phoenix_1988; @jacobs_generation_2014; @cross_american_1999]. Genocidal efforts to eradicate Native nations and cultures through the removal and separation of indigenous children persisted with explicit support from federal and state governments and leading social work organizations until the passage of the landmark Indian Child Welfare Act (ICWA) of 1978. ICWA provided tribal governments critical tools to exercise tribal sovereignty in US family courts and provides a critical framework for protecting Native families and respecting tribal sovereignty [@pevar_rights_2012; @jacobs_generation_2014]. Despite these protections, family separation continues to jeopardize and threaten the autonomy and wellbeing of Native families, and as a result, the very existence of tribal communities [@beardall_abolition_2021].

In this study, we describe the scale of the ongoing crisis of family separation by 1) identifying geographic patterns in the exposure of Native children and families to child welfare system contact and 2) identifying the stages in child welfare case processing where inequality is accelerated for Native children and families. This study provides new empirical evidence about where and how American Indian and Alaska Native families are being pushed into the child welfare system, and situates these findings in the long historical legacy of Native family separation.s

@@

@@

@@

@@

@@

@@

```{r nat-inv-age, fig.cap = "Age specific risk of experiencing first CPS investigation by state, 2014 - 2018 risk levels"}
ggplot(investigation_tables %>% 
         group_by(state, race_ethn, age) %>% 
         summarise(qmax = max(q), qmin=min(q), q = mean(q)),
       aes(x = age, y = q, ymin = qmin, ymax = qmax,
           color = race_ethn, fill = race_ethn)) + 
  geom_line() + 
  #geom_ribbon(alpha = 0.5, color = NA) + 
  facet_geo(~state) + 
  labs(color = "", fill = "") + 
  theme_minimal() + 
  labs(x = "Age", y = "Probability") + 
  theme(strip.text.x = element_text(size = 7),
        axis.text.x = element_text(size = 6),
        axis.text.y = element_text(size = 4)) + 
  scale_y_continuous(labels = scales::number_format(accuracy = 0.01)) + 
  scale_fill_brewer(palette = "Dark2")+
  scale_color_brewer(palette = "Dark2")

@@

```{r nat-fc-age, fig.cap = "Age specific risk of entering foster care, 2014 - 2018 risk levels"}
ggplot(fc_tables %>% 
         group_by(state, race_ethn, age) %>% 
         summarise(qmax = max(q), qmin=min(q), q = mean(q)),
       aes(x = age, y = q, ymin = qmin, ymax = qmax,
           color = race_ethn, fill = race_ethn)) + 
  geom_line() + 
  #geom_ribbon(alpha = 0.5, color = NA) + 
  facet_geo(~state) + 
  labs(color = "", fill = "") + 
  theme_minimal() + 
  labs(x = "Age", y = "Probability") + 
  theme(strip.text.x = element_text(size = 7),
        axis.text.x = element_text(size = 6),
        axis.text.y = element_text(size = 4)) + 
  scale_y_continuous(labels = scales::number_format(accuracy = 0.01)) + 
  scale_fill_brewer(palette = "Dark2")+
  scale_color_brewer(palette = "Dark2")
```

@@

ggplot(ncands_sub_inv_tab %>% 
  group_by(age, race_ethn, state) %>% 
    summarise(qmin = min(sub_inv),
              qmax = max(sub_inv),
              q = mean(sub_inv)),
  aes(x = age, y = q, color = race_ethn,
      ymin = qmin, ymax = qmax, fill = race_ethn)) +
  geom_line() + 
  # geom_ribbon(alpha = 0.5,
  #             color = NA) + 
  labs(fill = "", color = "",
       y = "Probability", x = "Age") + 
  facet_geo(~state) +
  scale_fill_brewer(palette = "Dark2")+
  scale_color_brewer(palette = "Dark2")+ 
  theme(axis.text.y = element_text(size = 5),
        axis.text.x = element_text(size = 5))
```

@@

ggplot(fc_cond_inv_tabs %>% 
         group_by(age, race_ethn, state) %>% 
         summarise(qmin = min(fc_cond_inv),
                   qmax = max(fc_cond_inv),
                   q = mean(fc_cond_inv)) %>% 
         filter(!(is.na(q)),
                q<1),
       aes(x = age, y = q, color = race_ethn,
           ymin = qmin, ymax = qmax, fill = race_ethn)) +
  geom_line() + 
  #geom_ribbon(alpha = 0.5,
  #            color = NA) + 
  labs(fill = "", color = "",
       y = "Probability", x = "Age") + 
  facet_geo(~state) +
  scale_fill_brewer(palette = "Dark2")+
  scale_color_brewer(palette = "Dark2")+ 
  theme(axis.text.y = element_text(size = 5),
        axis.text.x = element_text(size = 5))
```

@@

ggplot(fc_cond_sub_tabs %>% 
         group_by(age, race_ethn, state) %>% 
         summarise(qmin = min(fc_cond_sub),
                   qmax = max(fc_cond_sub),
                   q = mean(fc_cond_sub)) %>% 
         filter(!(is.na(q)),
                q<1),
       aes(x = age, y = q, color = race_ethn,
           ymin = qmin, ymax = qmax, fill = race_ethn)) +
  geom_line() + 
  # geom_ribbon(alpha = 0.5,
  #             color = NA) + 
  labs(fill = "", color = "",
       y = "Probability", x = "Age") + 
  facet_geo(~state) +
  scale_fill_brewer(palette = "Dark2")+
  scale_color_brewer(palette = "Dark2")+ 
  theme(axis.text.y = element_text(size = 5),
        axis.text.x = element_text(size = 5)) + 
  coord_cartesian(ylim = c(0,1))
```

@@

ggplot(non_icwa_cond_fc %>% 
         filter(race_ethn=="AIAN") %>% 
         group_by(age, state) %>% 
         summarise(qmin = min(non_icwa_fc),
                   qmax = max(non_icwa_fc),
                   q = mean(non_icwa_fc)) %>% 
         filter(!(is.na(q)),
                q<=1, qmax<=1),
       aes(x = age, y = q,
           ymin = qmin, ymax = qmax)) + 
  geom_line() +
  #geom_ribbon(alpha = 0.5, color = NA) + 
  facet_geo(~state) + 
  labs(x = "Age", y = "Proportion") + 
  theme(axis.text.y = element_text(size = 5),
        axis.text.x = element_text(size = 5))
```

@@

We examine data on the contemporary contact of AIAN children with the child welfare system at five distinct stages of case processing: investigations, substantiations, removals, placement settings, and termination of parental rights. Our findings reveal similar levels of national risk for AIAN children documented in other studies [@wildeman_cumulative_2020; @yi_cumulative_2020]. We find that AIAN children are more likely than white children to be placed into foster care, even when rates of investigation and substantiation of cases are similar to those for white children. We find that inequalities in case outcomes vary dramatically across regions of the country. Inequalities for AIAN children are typically low in the South and Northeast, and typically very high in the Midwest, Great Plains, Mountain West, and Northwest. While some states, such as Alaska, Minnesota, and North and South Dakota exhibit high inequalities in risks of investigation for AIAN children, we find that inequalities in family separation are typically a function of later case processing decisions. In many places, decisions made once a child is in the system, are critical in exacerbating the unequal exposure of AIAN children to family separation.
